Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-21 ind_files/04_field_eqns.tex
Minor rephrasing

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
We will assume that our process is nice enough that the hypotheses of the Radon-Nikodym theorem are satisfied and we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$. In one dimension, this is not necessarily a very big ask~\citep{walsh_introduction_1986,champagnat_unifying_2006,champagnat_individual_2008,rogers_demographic_2012}.

We can now use the same trick as in chapter \ref{chap_BD} and obtain a master equation by counting inflow and outflow of states. Any change to a state must be through the addition or subtraction of a single Dirac delta mass. For any state $\nu \in \mathcal{M}(\mathcal{T})$, the transition rate from $\nu - \delta_{x}$ to $\nu$ is simply $\mathcal{E}^{-}_{x}b(x|\nu)$, and similarly, the transition rate from $\nu+\delta_{x}$ to $\nu$ is $\mathcal{E}^{+}_{x}d(x|\nu)$. The transition rate out of $\nu$ to a state $\nu+\delta_x$ is just $b(x|\nu)$, and transition out to a state $\nu - \delta_x$ is just $d(x|\nu)$. Thus, integrating over all possible $x$ to obtain the total inflow and outflow rate for a state $\nu$, we see that $P(\nu,t)$ must satisfy:
\begin{equation}
\label{unnormalized_M_equation}
    \frac{\partial P}{\partial t}(\nu,t) = \int\limits_{\mathcal{T}}\left[(\mathcal{E}^{-}_{x}-1)b(x|\nu)P(\nu,t) + (\mathcal{E}^{+}_{x}-1)d(x|\nu)P(\nu,t)\right]dx
\end{equation}
This is the `Master equation' of our process.

>>>>

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
We will assume that our process is nice enough that the hypotheses of the Radon-Nikodym theorem are satisfied and we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$. In one dimension, this is not necessarily a very big ask~\citep{dawson_stochastic_1975,walsh_introduction_1986,etheridge_mathematical_2011,rogers_demographic_2012}.

We can now use the same trick as in chapter \ref{chap_BD} and obtain a master equation by counting inflow and outflow of states. Any change to a state must be through the addition or subtraction of a single individual (now a single Dirac mass). For any state $\nu \in \mathcal{M}(\mathcal{T})$, the transition rate from $\nu - \delta_{x}$ to $\nu$ is simply $\mathcal{E}^{-}_{x}b(x|\nu)$, and similarly, the transition rate from $\nu+\delta_{x}$ to $\nu$ is $\mathcal{E}^{+}_{x}d(x|\nu)$. The transition rate out of $\nu$ to a state $\nu+\delta_x$ is just $b(x|\nu)$, and transition out to a state $\nu - \delta_x$ is just $d(x|\nu)$. Thus, integrating over all possible $x$ to obtain the total inflow and outflow rate for a state $\nu$, we see that $P(\nu,t)$ must satisfy:
\begin{equation}
\label{unnormalized_M_equation}
    \frac{\partial P}{\partial t}(\nu,t) = \int\limits_{\mathcal{T}}\left[(\mathcal{E}^{-}_{x}-1)b(x|\nu)P(\nu,t) + (\mathcal{E}^{+}_{x}-1)d(x|\nu)P(\nu,t)\right]dx
\end{equation}
This is the `Master equation' of our process.

### 2023-03-21 ind_files/05_unification_discussion.tex
Minor rephrasing

In chapter \ref{chap_infD_processes}, I formulated a `field equation' for the evolution of one-dimensional quantitative traits in populations. To recap, given a one-dimensional quantitative trait that takes values in a trait space $\mathcal{T} \subseteq \mathbb{R}$, I defined the set $\mathcal{M}(\mathcal{T}) =  \left\{\sum_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}$, where each $\delta_{x_i}$ is a Dirac mass centered at $x_i$. I then formulated a model in which the population at time $t$ is characterized as a whole by a randomly varying density distribution (`stochastic field') $\nu^{(t)}: \mathcal{M}(\mathcal{T}) \to [0,\infty)$ such that for any subset $A \subseteq \mathcal{T}$, the number of individuals that have trait value in $A$ is given by integrating $\nu^{(t)}$ over $A$. The change of this field is determined entirely by two functionals, $b(x|\nu)$ and $d(x|\nu)$ from $\mathcal{T} \times \mathcal{M}(\mathcal{T})$ to $[0,\infty)$, that respectively describe the birth rate and death rate of individuals of type $x$ in a population $\nu$. Under the assumption that there exists a suitable system size parameter $K > 0$, I moved from the space of `number' distribution functions $\mathcal{M}(
\mathcal{T})$ to the space of `density' distribution functions $\mathcal{M}_{K}(\mathcal{T}) =  \left\{\sum_{i=1}^{n}\delta_{x_i}/K \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}$ via a functional analog of the system size expansion. By appropriately rescaling the birth and death rate functionals, I then determined that $P(\phi,t)$, the probability that the population is described by the distribution $\phi \in \mathcal{M}_{K}(\mathcal{T})$ at time $t$, (approximately) satisfies the `stochastic field equation':
\begin{equation}
\label{disc_functional_field_eqns}
\resizebox{\textwidth}{!}{$\displaystyle
\frac{\partial P}{\partial t}(\phi,t) = \int\limits_{\mathcal{T}}\left[-
\frac{\delta}{\delta\phi(x)} \big{\{} \big{[}\phi(x) w(x|\phi) + \mu Q(x|\phi)\big{]} P(\phi,t) \big{\}} + \frac{1}{2K}\frac{\delta^2}{\delta\phi(x)^2}\big{\{}\big{[}\phi(x) \tau(x|\phi) + \mu Q(x|\phi)\big{]}P(\phi,t)\big{\}}\right]dx$}
\end{equation}
where $w(x|\phi)$, $\tau(x|\phi)$, and $Q(x|\phi)$ are functionals that respectively describe the Malthusian fitness, per-capita turnover rate, and birth rate due to mutations (with mutation rate $\mu$) of type $x \in \mathcal{T}$ individuals (now a continuous variable) in a population $\phi$. We then saw that this equation yields some well-known frameworks of quantitative genetics in the infinite population ($K \to \infty$) limit, thus illustrating consistency with known theories.

>>>>

In chapter \ref{chap_infD_processes}, I formulated a `field equation' for the evolution of one-dimensional quantitative traits in populations. To recap, given a one-dimensional quantitative trait that takes values in a trait space $\mathcal{T} \subseteq \mathbb{R}$, I defined the set $\mathcal{M}(\mathcal{T}) =  \left\{\sum_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}$, where each $\delta_{x_i}$ is a Dirac mass centered at $x_i$. I then formulated a model for the evolutionary dynamics of of finite fluctuating populations of individuals that vary in the relevant quantitative trait as an $\mathcal{M}(\mathcal{T})$ valued birth-death process. The population at time $t$ is characterized by a randomly varying density distribution (`stochastic field') $\nu^{(t)} \in  \mathcal{M}(\mathcal{T})$ such that for any subset $A \subseteq \mathcal{T}$, the number of individuals that have trait value in $A$ is given by integrating $\nu^{(t)}$ over $A$. The change of this field is determined entirely by two functionals, $b(x|\nu)$ and $d(x|\nu)$ from $\mathcal{T} \times \mathcal{M}(\mathcal{T})$ to $[0,\infty)$, that respectively describe the birth rate and death rate of individuals of type $x$ in a population $\nu$. Under the assumption that there exists a suitable system size parameter $K > 0$, I moved from the space of `number' distribution functions $\mathcal{M}(
\mathcal{T})$ to the space of `density' distribution functions $\mathcal{M}_{K}(\mathcal{T}) =  \left\{\sum_{i=1}^{n}\delta_{x_i}/K \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}$ via a functional analog of the system size expansion. By appropriately rescaling the birth and death rate functionals, I then determined that $P(\phi,t)$, the probability that the population is described by the distribution $\phi \in \mathcal{M}_{K}(\mathcal{T})$ at time $t$, (approximately) satisfies the `stochastic field equation':
\begin{equation}
\label{disc_functional_field_eqns}
\resizebox{\textwidth}{!}{$\displaystyle
\frac{\partial P}{\partial t}(\phi,t) = \int\limits_{\mathcal{T}}\left[-
\frac{\delta}{\delta\phi(x)} \big{\{} \big{[}\phi(x) w(x|\phi) + \mu Q(x|\phi)\big{]} P(\phi,t) \big{\}} + \frac{1}{2K}\frac{\delta^2}{\delta\phi(x)^2}\big{\{}\big{[}\phi(x) \tau(x|\phi) + \mu Q(x|\phi)\big{]}P(\phi,t)\big{\}}\right]dx$}
\end{equation}
where $w(x|\phi)$, $\tau(x|\phi)$, and $Q(x|\phi)$ are functionals that respectively describe the Malthusian fitness, per-capita turnover rate, and birth rate due to mutations (with mutation rate $\mu$) of type $x \in \mathcal{T}$ individuals (now a continuous variable) in a population $\phi$. We then saw that this equation yields some well-known frameworks of quantitative genetics in the infinite population ($K \to \infty$) limit, thus illustrating consistency with known theories.
